---
title: "Introduction to SigsPack"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to SigsPack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Sigspack provides tools for easy computation of signature exposures from 
mutational catalogues. 

The package provides several features, allowing to read 
the primary mutation data, normalise the mutational catalogues if necessary and
compute signature exposures with their bootstrapped variation estimates.


## Loading the package

Download the package from Bioconductor

```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install("SigsPack")
```
        
Or install the development version of the package from Github

```{r, eval = FALSE}
devtools::install_github("bihealth/SigsPack")
```
        
And then load the package into your R session.

```{r Load, message=FALSE}
library(SigsPack)
```

## Loading a VCF

In most cases you will want to analyse real data, e.g. fit a sample's mutational
profile to known mutational signatures. You can easily load your data and use it
with the package.

```{r eval = FALSE}
if (require(BSgenome.Hsapiens.UCSC.hg19)) {
  sample <- vcf2mut_cat('test.vcf', BSgenome.Hsapiens.UCSC.hg19)
}
```

It will be transformed into a 'mutational profile' sorting all single nucleotide
variants according to their trinucleotide context and mutation. The result is a 
96 by 1 matrix that can be used as input to the analysis functions of the
package as well as in most other packages of this field.

## Simulating data

Instead of using real-life data it is also possible to generate 
synthetic data that can be used to analyse signatures stability. 
The following code generates 10 mutational catalogues  with exposure to the 
COSMIC signatures 2, 6, 15 and 27. The catalogues consist of mutation counts 
sampled from a distribution that is a linear combination of the aforementioned 
signatures. The weight of each of these signatures can optionally be specified, 
too. For convenience, the COSMIC reference signature matrix is included in the
package and used as default in the functions. However, it is also possible to 
use all functions with a custom signature matrix.
```{r}
data("cosmicSigs")

cats <- create_mut_catalogues(10, 500, P=cosmicSigs, sig_set = c(2,6,15,27))
knitr::kable(head(cats[['catalogues']]))

```


## Estimating signature exposures and bootstrapping samples

The function bootstrap_mut_catalogues bootstraps a sample returning a specified
amount of re-samples which can be used to gain a better variability estimation 
of the sample’s signature exposure.

The signature exposure is calculated using quadratic programming. This can be 
done on one or several samples at once. The COSMIC signatures have been included
in the package, and are used by default. However, it is possible to use your own
signature matrix, or use a sub-set of COSMIC signatures, instead of the whole 
matrix.
```{r}

reps <- bootstrap_mut_catalogues(n = 1000, original = cats[["catalogues"]][,1])

# using only signatures 4, 17, 23 and 30 for signature estimation
sigs <- signature_exposure(reps, sig_set = c(4,17,23,30))

print(sigs$exposures[,1])
```

With one command you can bootstrap a mutational catalogue and obtain a table and
a plot illustrating the results of the signature estimation for this sample and 
the bootstrapped re-samples. The plot shows the distribution of estimated 
signature exposure for all the re-samples, highlighting the one of the original 
mutational catalogue, thus providing insights on the reliability of the 
estimates.
```{r, fig.show='hold', fig.height=7, fig.width=7}

report <- summarize_exposures(reps[,1])

knitr::kable(
  head(report)
  )

```

## Tri-nucleotide contexts and normalization
Accurate exposures estimation requires matching tri-nucleotide frequencies 
between observations and the signature matrix. The COSMIC signature matrix 
provided in the package is relative to the whole genome (GRCh37) tri-nucleotide 
frequencies. So if you want to fit those signatures to exome data, the data need
to be normalized to match the signatures prior to exposures estimation. 

Let's say, that the vcf we derived the mutational catalogue from contained exome
data. In the case of this example, we want to scale our exome data to the
genome 'space' to match the COSMIC reference signatures. Hence, we need both the 
tri-nucleotide distribution of the human genome (GRCh37) and the exome that our 
data were collected on.

Extract the tri-nucleotide context frequencies from a genome (BSgenome 
object) or exome and use them to normalize the data.
```{r, eval = FALSE}
if (require(BSgenome.Hsapiens.UCSC.hg19)){
  genome_contexts <- get_context_freq(BSgenome.Hsapiens.UCSC.hg19)
  exome_contexts <- get_context_freq(BSgenome.Hsapiens.UCSC.hg19, 'example_exome.bed')
  normalized_mut_cat <- normalize(sample, exome_contexts, hg19context_freq)
}

```


## sessionInfo
```{r}
sessionInfo()
```
